Type-Safe Modular Parsing

### 2016-12-18 paper/paper.tex
First comments on Section 2
Done 2.2, rewriting 2.3
work on the intro

\usepackage{microtype}%if unwanted, comment out or use option "draft"
\usepackage[table,xcdraw]{xcolor}
\usepackage{color}
\usepackage{amsmath}
\usepackage{stmaryrd}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{fancyvrb}
\usepackage{url}
\usepackage{pstricks,pst-node,pst-tree}
\usepackage{theorem}
%% \usepackage{mathpartir}
\usepackage{bbm}
\usepackage{pgf}
\usepackage{multirow}

@@

%{\theorembodyfont{\sffamily} \newtheorem{theorem}{Theorem}}
%{\theorembodyfont{\sffamily} \newtheorem{lemma}{Lemma}}
%\newtheorem{theorem}{Theorem}
%\newtheorem{lemma}{Lemma}
%\newenvironment{proof}{\textbf{Proof:\hspace{4mm}}}{$\Box$}
\newcommand{\authornote}[3]{{\color{#2} {\sc #1}: #3}}
\newcommand\bruno[1]{\authornote{bruno}{red}{#1}}
\newcommand\yanlin[1]{\authornote{yanlin}{purple}{#1}}
%\newcommand{\hl}[1]{\textcolor{red}{#1}}
\newcommand\huang[1]{\authornote{huang}{blue}{#1}}
\newcommand\haoyuan[1]{\authornote{haoyuan}{cyan}{#1}}

@@

\newcommand\mixin{\textbf{@Mixin} }
\newcommand\name{\textbf{Parser}\xspace}
\usepackage{xspace}

@@

  This paper investigates solutions for the problem of \emph{modular
    parsing}. We focus on \emph{semantic} modularity and not just
  \emph{syntactic} modularity. That is, the solutions should not only
  allow complete parsers to be built out of modular parsing
  components, but also enable the parsing components to be \emph{modularly
  type-checked} and \emph{separately compiled}. We present \name: a parsing
  combinator library that enables modular parsing. The library is
  built on top of a Packrat parsing library, but adds new parsing
  combinators to enable modular parsing. The new parsing combinators 
  employ \emph{delegation-based} techniques and \emph{Object Algebras} 
  to support extensibility. The choice of Packrat parsing over other
  parsing techniques turns out to be important for achieving
  performance in a modular setting. To evaluate \name 
  we conduct a  case study based on the 
  ``Types and Programming Languages'' interpreters. The case study 
  shows that \name is effective at reusing parsing code from existing
  interpreters, and the total parsing code is 60\% shorter than the
  non-modular parsing code.

>>>>

\usepackage{microtype}%if unwanted, comment out or use option "draft"
\usepackage[table,xcdraw]{xcolor}
\usepackage{color}
\usepackage{amsmath}
\usepackage{stmaryrd}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{fancyvrb}
\usepackage{url}
\usepackage{pstricks,pst-node,pst-tree}
\usepackage{theorem}
\usepackage{syntax}
\usepackage{bbm}
\usepackage{pgf}
\usepackage{multirow}

@@

%{\theorembodyfont{\sffamily} \newtheorem{theorem}{Theorem}}
%{\theorembodyfont{\sffamily} \newtheorem{lemma}{Lemma}}
%\newtheorem{theorem}{Theorem}
%\newtheorem{lemma}{Lemma}
%\newenvironment{proof}{\textbf{Proof:\hspace{4mm}}}{$\Box$}
\newcommand\inlinecode[1]{\lstinline[language=PlainCode]{#1}}
\newcommand{\authornote}[3]{{\color{#2} {\sc #1}: #3}}
\newcommand\bruno[1]{\authornote{bruno}{red}{#1}}
\newcommand\yanlin[1]{\authornote{yanlin}{purple}{#1}}
%\newcommand{\hl}[1]{\textcolor{red}{#1}}
\newcommand\huang[1]{\authornote{huang}{blue}{#1}}
\newcommand\haoyuan[1]{\authornote{haoyuan}{cyan}{#1}}

@@

\newcommand\mixin{\textbf{@Mixin} }
\newcommand\name{\textbf{Parser}\xspace}
\usepackage{xspace}
\renewcommand{\paragraph}[1]{\vspace{2pt}\noindent {\bf #1}}

@@

  This paper investigates solutions for the problem of \emph{modular
    parsing}. We focus on \emph{semantic} modularity and not just
  \emph{syntactic} modularity. That is, the solutions should not only
  allow complete parsers to be built out of modular parsing
  components, but also enable the parsing components to be
  \emph{modularly type-checked} and \emph{separately compiled}. We
  present \name: a parsing combinator library that enables modular
  parsing. Interestingly, the modularity requirements for modular
  parsing rule out several existing parsing combinator approaches,
  which rely on some non-modular techniques. Fortunatelly Packrat
  parsing techniques, provide solutions for such modularity problems,
  and enable reasonable performance in a modular setting.  Thus \name
  is built on top of a Packrat parsing library, but adds new parsing
  combinators to enable modular parsing. The new parsing combinators
  employ \emph{delegation-based} techniques and \emph{Object Algebras}
  to support extensibility. To evaluate \name we conduct a case study
  based on the ``Types and Programming Languages'' interpreters. The
  case study shows that \name is effective at reusing parsing code
  from existing interpreters, and the total parsing code is 60\%
  shorter than the non-modular parsing code.

### 2016-12-18 paper/sections/Introduction.tex
work on the intro

The quest for improved modularity, variability and extensibility of
programs has been going on since the early days of Software
Engineering~\cite{}. Modern Programming Languages enable a certain
degree of modularity, but they have limitations as illustrated by
well-known problems such as the Expression Problem~\cite{}. To address
the limitations of Programming Languages several different approaches
have been proposed in the past. Existing approaches can be broadly
divided into two categories: \emph{syntactic} or \emph{semantic} 
modularization techniques. Syntactic modularization techniques enable 
the separate development of software components or features that can
later be combined into complete programs. However, very often, the 
composition of components in syntactic modularization techniques 
amounts to a form of textual composition of the code from multiple 
components/features. Semantic modularization
techniques go one step further, and also enable those components or
features to be modularly type-checked and separately compiled. This
enables the composition of compiled binaries, and also ensures the
type-safety of the composed code.

Many approaches use syntactic modularization techniques. Examples
include tools for developing Software-Product Lines~\cite{}, some
Language Workbenches~\cite{}, or extensible parser generators.  In
those approaches textual composition techniques such as
\emph{superimposition}~\cite{} enable the development . Syntactic
modularization techniques have also been applied to the problem of
\emph{extensible parsing}. There are several approaches that enable
the development of \emph{syntactically} modular parsers or
grammars. However these approaches do not support separate compilation
or modular type-checking that we strive for in this work.

\begin{comment}
 such as AHEAD or 
FeatureHouse, which enable the development of highly reusable 
software components that can be combined to form complete 
software products.
\end{comment}

There are also several approaches to semantic modularity. For example,
family polymorphism enables the development of highly reusable
components and can even be used to develop Software-Product Lines. 
... However, as far as we are aware, the problem of modular parsing 
has not been studied in semantic modularization
approaches.

  This paper investigates solutions for the problem of \emph{modular
    parsing}. We focus on \emph{semantic} modularity and not just
  \emph{syntactic} modularity. That is, the solutions should not only
  allow complete parsers to be built out of modular parsing
  components, but also enable the parsing components to be \emph{modularly
  type-checked} and \emph{separately compiled}. We present \name: a parsing
  combinator library that enables modular parsing. The library is
  built on top of a Packrat parsing library, but adds new parsing
  combinators to enable modular parsing. The new parsing combinators 
  employ \emph{delegation-based} techniques and \emph{Object Algebras} 
  to support extensibility. The choice of Packrat parsing over other
  parsing techniques turns out to be important for achieving
  performance in a modular setting. To evaluate \name 
  we conduct a  case study based on the 
  ``Types and Programming Languages'' interpreters. The case study 
  shows that \name is effective at reusing parsing code from existing
  interpreters, and the total parsing code is 60\% shorter than the
  non-modular parsing code.

>>>>

The quest for improved modularity, variability and extensibility of
programs has been going on since the early days of Software
Engineering~\cite{}. Modern Programming Languages enable a certain
degree of modularity, but they have limitations as illustrated by
well-known problems such as the Expression Problem~\cite{}. The
Expression Problem refers to the difficulty of writing data
abstractions that can be easily extended with both new operations and
new data variants. Traditionally the kinds of data abstraction found
in functional languages can be extended with new operations, but
adding new data variants is difficult. The traditional object-oriented
approach to data abstraction facilitates adding new data variants
(classes), while adding new operations is more difficult.

A reason why a solution to the Expression Problem is important in
practice is that it is necessary for the development of
\emph{Software-Product Lines} (SPLs)~\cite{}. A software-product line
is a reusable set of components, which can be combined in multiple ways
to obtain different programs. Programming languages offer a concrete
example for SPLs. A SPL for programming languages would allow us to
model various typical operations of programming languages (such as
evaluation, compilation, or parsing) for various different language
constructs (such as binding, arithmetic, conditional or loops)
independently and separately. For example, evaluation components could be defined
independently for binding and arithmetic constructs. If the language
to be implemented is the pure lambda calculus, only evaluation of
binding constructs is necessary. However, more realistic programming
languages will include arithmetic constructs, and will require 
evaluation for such constructs as well. In this case 
both the component for evaluation of binders and arithmetic 
expressions can be combined to implement the desired functionality.

\begin{comment}
Most programming languages share alot of features in
common.

For example, most languages have language constructs for:
binding (such as variables, functions, and function applications);
basic arithmetic operations; basic logic and conditional operations;
loops; as well as various other features. For each language construct,
various operations (such as evaluation, compilation, or parsing) need
to be implemented. It is reasonable to wonder whether we can simply
implement those features independently of a particular implementation
of a programming language. Evaluation could be defined independently 
for binding and arithmetic constructs. If the language to be
implemented is the pure lambda calculus, only evaluation of binding 
constructs is necessary. Thus only the component that implements 
evaluation for binding needs to be used in such an implementation.
However, more realistic programming languages 
will include arithmetic constructs, and will require an evaluation
function for those.

Then it would be possible to \emph{reuse}
some of those features in \emph{multiple} different implementations of
programming languages. Essentially, this would enable a SPL for
programming languages, where all

A solution to the Expression Problem could ena

A concrete 
example that illustrates this issue is 
\end{comment}

To address the modularity limitations of Programming Languages several
different approaches have been proposed in the past. Existing
approaches can be broadly divided into two categories:
\emph{syntactic} or \emph{semantic} modularization
techniques. Syntactic modularization techniques are quite popular in
practice. Examples include many tools for developing Software-Product
Lines~\cite{}, some Language Workbenches~\cite{}, or extensible parser
generators~\cite{}.  Most syntactic approaches employ textual
composition techniques such as \emph{superimposition}~\cite{} to
enable the development modular program features. Such textual
composition techniques collect the code for multiple features and
merge it together when a concrete combination of features is needed
for a particular program. The typical drawback of such techniques is
the lack of modular type-checking and separate compilation:
``\emph{most feature-oriented implementation mechanisms lack proper
  interfaces and support neither modular type checking nor separate
  compilation}''~\cite{kastner11road}\bruno{reference to ``The road to
  Feature Modularity''}. Syntactic modularization techniques have also
been applied to the problem of \emph{extensible parsing}. There are
several approaches~\cite{} that enable the development of
\emph{syntactically} modular parsers or grammars. However these
approaches do not support separate compilation or modular
type-checking either.\bruno{mention that a reason why syntactic 
modularization techniques are popular is simplicity (in implementation
and also use).}

Semantic modularization techniques move away
from syntactic composition techniques that rely on the textual source
code. This allows us to go one step further in terms of modularity,
and also enable components or features to be modularly type-checked
and separately compiled. Modular type-checking and separate
compilation are desirable properties to have from a Software
Engineering point-of-view, and enable the composition of compiled
binaries as well as ensuring the type-safety of the code composed of
multiple components. Examples of semantic modularization techniques 
include various approaches to \emph{family polymorphism}~\cite{}, as 
well as several techniques for solving the Expression Problem. 
\bruno{strenghten? mention dynamic languages (separate compilation 
but no modular type-checking), and techniques that weaken type-safety 
requirements.} Most semantic
modularization techniques have focused on operations that traverse or 
or process extensible datastructures, such as ASTs.
\bruno{Object Algebras here?} However, as far as
we know there is little work on operations that build/produce ASTs. 
In particular the problem of modular parsing has not been studied in
semantic modularization approaches. This is a shame because, to
realise the vision of software product-lines for programming
languages, modular parsing is a necessity.

\paragraph{Modular and Extensible Parsing}
  This paper investigates presents \name: a parsing
  combinator library that enables modular parsing.
  \name provides a solution for the problem of \emph{semantic modular
    parsing}. That is, the solutions should not only
  allow complete parsers to be built out of modular parsing
  components, but also enable the parsing components to be \emph{modularly
  type-checked} and \emph{separately compiled}. \name is
  built on top of a Packrat parsing library, but adds new parsing
  combinators to enable modular parsing. The new parsing combinators 
  employ \emph{delegation-based} techniques and \emph{Object Algebras} 
  to support extensibility. The choice of Packrat parsing over other
  parsing techniques turns out to be important for achieving
  performance in a modular setting. \bruno{left recursion, and
    backtracking removal.}

  By analising the \emph{full} grammar it is possible to remove
  backtracking, which would otherwise increase parsing times. Many
  parsing combinator libraries routinely use backtracting elimination
  to achieve performance. However, in a modular setting this technique
  cannot be used, because the full grammar is not known. Thus we have
  to be very conservative at eliminating backtracting. Unfortunatelly,
  this has a severe impact on performance.

  To evaluate \name we conduct a case study based on the ``Types and
  Programming Languages'' (TAPL) interpreters.  The case study shows
  that \name is effective at reusing parsing code from existing
  interpreters, and the total parsing code is 60\% shorter than the
  non-modular parsing code.\bruno{comment on the efficiency}

In summary our contributions are:

\begin{itemize}

\item {\name:} A parser combinator library that allows the development 
of modular parser.

\item {{\bf TAPL case study:}} We conduct a case study with 18 interpreters
  from the TAPL book. The case study show the effectiveness of modular 
  parsing in terms of reuse.

\end{itemize}
